A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-25 text/formal-defnition.tex
Fixing proofs.
Reflect comments from John.

The variable $x$ is bound in $M$ of $\lambda x:\tau.M$ and in $\sigma$
of $\Pi x:\tau.\sigma$, as usual; the stage variable $\alpha$ is
bound in $M$ of $\Lambda \alpha.M$ and $\tau$ of $\F\alpha.\tau$.
The notion of free variables is defined in a standard manner.
We write $\FV(M)$ and $\FTV(M)$ for the set of free variables and the set of free stage variables in $M$, respectively.  Similarly, $\FV(\tau)$, $\FTV(\tau)$,
$\FV(K)$, and $\FTV(K)$ are defined.
We sometimes abbreviate $\Pi x:\tau_1.\tau_2$ to $\tau_1 \rightarrow \tau_2$ if
$x$ is not a free variable of $\tau_2$.
% Free variables
We identify $\alpha$-convertible terms and assume the names of bound variables are pairwise distinct.
% \AI{This should be mentioned after free variables are introduced.}
% \AI{What about other binders such as $\Pi$?}

@@

  {\small
    \begin{center}
      \infrule[\textsc{QK-Csp}]{%
        \G\V K \E J @ A
      }{
        \G\V K \E J @ A\alpha
      }
      \hfil
      \infrule[\QTCsp]{
        \G\V \tau \E \sigma :: K@A
      }{
        \G\V \tau \E \sigma :: K@{A\alpha}
      }
      \\[2mm]
      \infrule[\QTApp]{%
        \G\V \tau \E \sigma :: (\Pi x:\rho.K)@A \andalso
        \G\V M \E N : \rho @A
      }{
        \G\V \tau\ M \E \sigma\ N :: K[x \mapsto M]@A
      }
    \end{center}
  }

@@

The helper function vadd$_1$ takes a stage, the length $n$ of vectors, and two quoted vectors and returns code that computes the addition of the given two vectors:
%
\newcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
\begin{tabbing}
	  $\textbf{let}\ \text{vadd}_1 : \F\alpha.\Pi n:\I.\TW_\alpha\Vpn\to\TW_\alpha\Vpn\to\TW_\alpha\Vpn$                                \\
	  \hspace{6mm} \= $= \textbf{fix}\ f.\Lambda\alpha.\lambda n:\I.\ \lambda v_1:\TW_\alpha\Vpn.\ \lambda v_2:\TW_\alpha\Vpn.$            \\
	  \> \hspace{6mm} \= $\textbf{if}\ n = 0 \ \textbf{then}\ \TB_\alpha \text{nil}$ \\
	  \>\> $\textbf{else}\ \TB_\alpha ($ \= $\textbf{let}\ t_1 = \text{tail}\ (\TBL_\alpha v_1)\ \textbf{in}$ \\
	  \>\>\> $\textbf{let}\ t_2 = \text{tail}\ (\TBL_\alpha v_2)\ \textbf{in}$ \\
          \>\>\> $\text{cons}\ (\text{head}\ (\TBL_\alpha v_1) + \text{head}\ (\TBL_\alpha v_2))\ (\TBL_\alpha f\ (n-1)\ t_1\ t_2))$
\end{tabbing}
(Here, we assume that the type system can know whether $n=0$ when
\textbf{then}- and \textbf{else}-branches are typechecked so that
both branches can be given type \(\TW_\alpha \text{Vector }n\).)

@@

The subscripts $A$ and $B$ in $E^A_B$ stand for the stage of the evaluation context and of the hole, respectively.  The grammar represents that staged reduction is left-to-right and call-by-value and terms under \(\Lambda\) is reduced.
A few examples of evaluation context are shown below:
\begin{align*}
  \square\ (\lambda x:\I.x)                  & \in  ECtx^\varepsilon_\varepsilon \\
  \Lambda\alpha.\square\ \epsilon            & \in ECtx^\varepsilon_\varepsilon  \\
  \TBL_\alpha \TB_\alpha \TBL_\alpha \square & \in ECtx^\alpha_\varepsilon
\end{align*}
%
We write $E^A_B[M]$ for the term obtained by filling the hole $\square$ in $E^A_B$ by $M$.

>>>>

The variable $x$ is bound in $M$ by $\lambda x:\tau.M$ and in $\sigma$
by $\Pi x:\tau.\sigma$, as usual; the stage variable $\alpha$ is
bound in $M$ by $\Lambda \alpha.M$ and $\tau$ by $\F\alpha.\tau$.
The notion of free variables is defined in a standard manner.
We write $\FV(M)$ and $\FTV(M)$ for the set of free variables and the set of free stage variables in $M$, respectively.  Similarly, $\FV(\tau)$, $\FTV(\tau)$,
$\FV(K)$, and $\FTV(K)$ are defined.
We sometimes abbreviate $\Pi x:\tau_1.\tau_2$ to $\tau_1 \rightarrow \tau_2$ if
$x$ is not a free variable of $\tau_2$.
% Free variables
We identify $\alpha$-convertible terms and assume the names of bound variables are pairwise distinct.
% \AI{This should be mentioned after free variables are introduced.}
% \AI{What about other binders such as $\Pi$?}

@@

  {\small
    \begin{center}
      \infrule[\textsc{QK-Csp}]{%
        \G\V K \E J @ A
      }{
        \G\V K \E J @ A\alpha
      }
      \hfil
      \infrule[\QTCsp]{
        \G\V \tau \E \sigma :: *@A
      }{
        \G\V \tau \E \sigma :: *@{A\alpha}
      }
      \\[2mm]
      \infrule[\QTApp]{%
        \G\V \tau \E \sigma :: (\Pi x:\rho.K)@A \andalso
        \G\V M \E N : \rho @A
      }{
        \G\V \tau\ M \E \sigma\ N :: K[x \mapsto M]@A
      }
    \end{center}
  }

@@

The helper function vadd$_1$ takes a stage, the length $n$ of vectors, and two quoted vectors and returns code that computes the addition of the given two vectors:
%
\newcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
\begin{tabbing}
	  $\textbf{let}\ \text{vadd}_1 : \F\alpha.\Pi n:\I.\TW_\alpha\Vpn\to\TW_\alpha\Vpn\to\TW_\alpha\Vpn$                                \\
	  \hspace{6mm} \= $= \textbf{fix}\ f.\Lambda\alpha.\lambda n:\I.\ \lambda v_1:\TW_\alpha\Vpn.\ \lambda v_2:\TW_\alpha\Vpn.$            \\
	  \> \hspace{6mm} \= $\textbf{if}\ n = 0 \ \textbf{then}\ \TB_\alpha \text{nil}$ \\
	  \>\> $\textbf{else}\ \TB_\alpha ($ \= $\textbf{let}\ t_1 = \text{tail}\ (\TBL_\alpha v_1)\ \textbf{in}$ \\
	  \>\>\> $\textbf{let}\ t_2 = \text{tail}\ (\TBL_\alpha v_2)\ \textbf{in}$ \\
          \>\>\> $\text{cons}\ (\text{head}\ (\TBL_\alpha v_1) + \text{head}\ (\TBL_\alpha v_2))\ (\TBL_\alpha f\ (n-1)\ t_1\ t_2))$
\end{tabbing}
(Here, we assume that the type system can determine whether $n=0$ when
\textbf{then}- and \textbf{else}-branches are typechecked so that
both branches can be given type \(\TW_\alpha \text{Vector }n\).)

@@

The subscripts $A$ and $B$ in $E^A_B$ stand for the stage of the evaluation context and of the hole, respectively.  The grammar represents that staged reduction is left-to-right and call-by-value and terms under \(\Lambda\) are reduced.
A few examples of evaluation context are shown below:
\begin{align*}
  \square\ (\lambda x:\I.x)                  & \in  ECtx^\varepsilon_\varepsilon \\
  \Lambda\alpha.\square\ \epsilon            & \in ECtx^\varepsilon_\varepsilon  \\
  \TBL_\alpha \TB_\alpha \TBL_\alpha \square & \in ECtx^\alpha_\varepsilon
\end{align*}
%
We write $E^A_B[M]$ for the term obtained by filling the hole $\square$ in $E^A_B$ by $M$.

### 2019-06-25 text/informal-overview.tex
Reflect comments from John.

In \LTP, brackets (quasi-quotation) and escape (unquote) are written
$\TB_\alpha M$ and $\TBL_\alpha M$, respectively.  For example,
$\TB_\alpha (1 + 1)$ represents code of expression $1 + 1$ and thus
evaluates to itself.  Escape $\TBL_\alpha M$ may appear under
$\TB_\alpha$; it evaluates $M$ to a code value and splice it into the
surrounding code fragment.  Such splicing is expressed by the
following reduction rule:
\begin{align*}
	\TBL_\alpha (\TB_\alpha M) \longrightarrow M .
\end{align*}
% Using the type constructor $\TW_\alpha$ for code values, the type of
% $\TB_\alpha M$ is given $\TW_\alpha \tau$ if $M$ has type $\tau$; the type
% of $\TBL_\alpha M$ is $\tau$ if $M$ has $\TW_\alpha \tau$.

The subscript $\alpha$ in $\TB_\alpha$ and $\TBL_\alpha$ is a \textit{stage
  variable}\footnote{%
  In Hanada and Igarashi~\cite{Hanada2014}, it was called a
  \textit{transition variable}, which is derived from correspondence
  to modal logic, studied by Tsukada and Igarashi~\cite{Tsukada}.} and
a sequence of stage variables is called \textit{stage}.  Intuitively,
they represent the depth of nested brackets.  Stage variables can be
abstracted by $\Lambda\alpha.M$ and instantiated by an application
$M\ A$ to stages.  For example,
$\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10)\ 5)$ is a code value,
where \(\alpha\) is abstracted.  If it is applied to \(\alpha_1 \cdots \alpha_n\), \(\TB_\alpha\) becomes \(\TB_{\alpha_1} \cdots \TB_{\alpha_n}\); in particular,
if \(n = 0\), \(\TB_\alpha\) disappears.  So, an
application of $\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10)\ 5)$
to the empty sequence \(\varepsilon\) reduces to
(unquoted) \((\lambda x:\I.x+10)\ 5\) and to 15.  In other words, application of of
\(\Lambda\)-abstraction to $\varepsilon$ corresponds to \texttt{run}.
% If \(\alpha\) is instantiated with, say, \(\beta\gamma\), then
% \(\TB_\alpha\) becomes nested brackets \(\TB_{\beta} \TB_\gamma\).
This is expressed by the following reduction rule:
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
where stage substitution \([\alpha \mapsto A]\) manipulates nesting of
\(\TB_\alpha\) and \(\TBL_\alpha\) (and also \(\%_\alpha\) as we see later).

@@

Second, an indexed type such as Vector\ 3 or Vector\ $x$ is well
formed only at the stage(s) where the index term is well typed.  Since
constant \(3\) is well typed at every stage (if it is declared in the
signature), so Vector\ 3 is a well-formed type at every stage, too.
However, Vector\ $M$ is well formed only at the stage where index term
$M$ is typed.  Thus, the kinding judgment
of \LMD takes the form \(\G\V \tau :: K @ A\), where stage $A$ stands for
where \(\tau\) is well formed.  For example,
given \(\text{Vector}:: \I \rightarrow *\) in the signature \(\Sigma\),
\(x:\I@\varepsilon \V \text{Vector }x :: * @\varepsilon\) can be
derived but
neither
\(x:\I@\alpha \V \text{Vector }x :: * @\varepsilon\)
nor 
\(x:\I@\epsilon \V \text{Vector }x :: * @\alpha\)
can be.

>>>>

In \LTP, brackets (quasi-quotation) and escape (unquote) are written
$\TB_\alpha M$ and $\TBL_\alpha M$, respectively.  For example,
$\TB_\alpha (1 + 1)$ represents code of expression $1 + 1$ and thus
evaluates to itself.  Escape $\TBL_\alpha M$ may appear under
$\TB_\alpha$; it evaluates $M$ to a code value and splices it into the
surrounding code fragment.  Such splicing is expressed by the
following reduction rule:
\begin{align*}
	\TBL_\alpha (\TB_\alpha M) \longrightarrow M .
\end{align*}
% Using the type constructor $\TW_\alpha$ for code values, the type of
% $\TB_\alpha M$ is given $\TW_\alpha \tau$ if $M$ has type $\tau$; the type
% of $\TBL_\alpha M$ is $\tau$ if $M$ has $\TW_\alpha \tau$.

The subscript $\alpha$ in $\TB_\alpha$ and $\TBL_\alpha$ is a \textit{stage
  variable}\footnote{%
  In Hanada and Igarashi~\cite{Hanada2014}, it was called a
  \textit{transition variable}, which is derived from correspondence
  to modal logic, studied by Tsukada and Igarashi~\cite{Tsukada}.} and
a sequence of stage variables is called a \textit{stage}.  Intuitively,
they represent the depth of nested brackets.  Stage variables can be
abstracted by $\Lambda\alpha.M$ and instantiated by an application
$M\ A$ to stages.  For example,
$\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10)\ 5)$ is a code value,
where \(\alpha\) is abstracted.  If it is applied to \(\alpha_1 \cdots \alpha_n\), \(\TB_\alpha\) becomes \(\TB_{\alpha_1} \cdots \TB_{\alpha_n}\); in particular,
if \(n = 0\), \(\TB_\alpha\) disappears.  So, an
application of $\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10)\ 5)$
to the empty sequence \(\varepsilon\) reduces to
(unquoted) \((\lambda x:\I.x+10)\ 5\) and to 15.  In other words, application of of
\(\Lambda\)-abstraction to $\varepsilon$ corresponds to \texttt{run}.
% If \(\alpha\) is instantiated with, say, \(\beta\gamma\), then
% \(\TB_\alpha\) becomes nested brackets \(\TB_{\beta} \TB_\gamma\).
This is expressed by the following reduction rule:
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
where stage substitution \([\alpha \mapsto A]\) manipulates nesting of
\(\TB_\alpha\) and \(\TBL_\alpha\) (and also \(\%_\alpha\) as we see later).

@@

Second, an indexed type such as Vector\ 3 or Vector\ $x$ is well
formed only at the stage(s) where the index term is well typed.  Since
constant \(3\) is well typed at every stage (if it is declared in the
signature), Vector\ 3 is a well-formed type at every stage, too.
However, Vector\ $M$ is well formed only at the stage where index term
$M$ is typed.  Thus, the kinding judgment
of \LMD takes the form \(\G\V \tau :: K @ A\), where stage $A$ stands for
where \(\tau\) is well formed.  For example,
given \(\text{Vector}:: \I \rightarrow *\) in the signature \(\Sigma\),
\(x:\I@\varepsilon \V \text{Vector }x :: * @\varepsilon\) can be
derived but
neither
\(x:\I@\alpha \V \text{Vector }x :: * @\varepsilon\)
nor 
\(x:\I@\epsilon \V \text{Vector }x :: * @\alpha\)
can be.

### 2019-06-25 text/introduction.tex
Reflect comments from John.

Multi-stage programming makes it easier for programmers to implement
generation and execution of code at run time by providing language
constructs for composing and running pieces of code as first-class
values.  A promising application of multi-stage programming is
(run-time) code specialization, which is to generate program code
specialized to partial inputs to the program and such applications are
studied in the literature~\cite{8384206,taha2007gentle}  
% \AI{We should list applications of multi-stage programming, not necessarily of MetaOCaml.}

@@

\subsection{Our Work}
In this paper, we develop a new multi-stage calculus \LMD by extending
the existing multi-stage calculus \LTP\cite{Hanada2014} with dependent
types and study its properties.  We base our work on \LTP, in which
the four multi-stage constructs are handled slightly differently from
MetaOCaml, because its type system and semantics are arguably simpler
than \(\lambda^\alpha\)~\cite{taha2003environment}, which formalizes
the design of MetaOCaml more faithfully.  Dependent types are based on
\LLF~\cite{attapl}, which has one of the simplest form of dependent
types.  Our technical contribution are summarized as follows:
\begin{itemize}
\item We give a formal definition of \LMD with its syntax, 
  type system and two kinds of reduction---full and staged reduction.
\item We show preservation, strong normalization, and confluence for
  full reduction; and show unique decomposition (and progress as its
  corollary) for staged reduction, which formalizes program execution.
\end{itemize}
The combination of multi-stage programming and dependent types has
been discussed by Brady and Hammond~\cite{brady2006dependently} but,
to our knowledge, our work is a first formal calculus of dependently
typed multi-stage programming.

>>>>

Multi-stage programming makes it easier for programmers to implement
generation and execution of code at run time by providing language
constructs for composing and running pieces of code as first-class
values.  A promising application of multi-stage programming is
(run-time) code specialization, which generates program code
specialized to partial inputs to the program and such applications are
studied in the literature~\cite{8384206,mainland2012metahaskell,taha2007gentle}.
% \AI{We should list applications of multi-stage programming, not necessarily of MetaOCaml.}

@@

\subsection{Our Work}
In this paper, we develop a new multi-stage calculus \LMD by extending
the existing multi-stage calculus \LTP\cite{Hanada2014} with dependent
types and study its properties.  We base our work on \LTP, in which
the four multi-stage constructs are handled slightly differently from
MetaOCaml, because its type system and semantics are arguably simpler
than \(\lambda^\alpha\)~\cite{taha2003environment}, which formalizes
the design of MetaOCaml more faithfully.  Dependent types are based on
\LLF~\cite{attapl}, which has one of the simplest form of dependent
types.  Our technical contributions are summarized as follows:
\begin{itemize}
\item We give a formal definition of \LMD with its syntax, 
  type system and two kinds of reduction---full and staged reduction.
\item We show preservation, strong normalization, and confluence for
  full reduction; and show unique decomposition (and progress as its
  corollary) for staged reduction, which formalizes program execution.
\end{itemize}
The combination of multi-stage programming and dependent types has
been discussed by Brady and Hammond~\cite{brady2006dependently} but,
to our knowledge, our work is a first formal calculus of dependently
typed multi-stage programming.
